\LaTeXe\ guide for the Journal of Functional Programming

### 2019-04-26 papers/jfp/3-abstractions.tex
Expand S8.1: Partial stores and exceptions
Tweaks to S3

This section presents purely functional abstractions that allow us to express
all the intricacies of build systems discussed in~\S\ref{sec-background},
and design complex build systems from simple primitives. Specifically, we
present the \emph{task} and \emph{build} abstractions in~\S\ref{sec-task}
and~\S\ref{sec-general-build}, respectively. Sections~\S\ref{sec-scheduler},
\S\ref{sec-rebuilder} and~\S\ref{sec-implementations} scrutinise the
abstractions further and provide concrete implementations for several build
systems.

@@

We will model build systems concretely, as Haskell programs. To that end,
Fig.~\ref{fig-types} provides the type signatures for all key abstractions
introduced in the paper. For example, \hs{Store}~\hs{i}~\hs{k}~\hs{v} is the
type of stores, with several associate functions (\hs{getValue}, etc.). We use
\hs{k} as a type variable ranging over keys, \hs{v} for values, and \hs{i} for
the persistent build information. Fig.~\ref{fig-stdlib} lists standard library
definitions.

@@

\noindent
Here \hs{c} stands for \emph{constraint}, such as \hs{Applicative}
(\S\ref{sec-why-polymorphism} explains why we need it).
A \hs{Task} describes a single build task, while \hs{Tasks} associates a
\hs{Task} to every non-input key; input keys are associated with \hs{Nothing}.
The highly-abstracted type \hs{Task} describes how to build a value when given
a way to build its dependencies, and is best explained by an example.
Consider this \Excel spreadsheet:
% \footnote{Readers familiar with \emph{lenses} or \emph{profunctor optics} might
% recognise a familiar pattern. We discuss this in~\S\ref{sec-related-optics}.}

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Functor     f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard State monad from Control.Monad.State
data State s a
instance Monad (State s)
get       :: State s s
gets      :: (s -> a) -> State s a
put       :: s -> State s ()
modify    :: (s -> s) -> State s ()
runState  :: State s a -> s -> (a, s)
execState :: State s a -> s -> s
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard types from Data.Functor.Identity and Data.Functor.Const
newtype Identity a = Identity { runIdentity :: a }
newtype Const m a  = Const    { getConst    :: m }
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
instance Functor (Const m) where
    fmap _ (Const m) = Const m
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
instance Monoid m => Applicative (Const m) where
    pure _              = Const mempty   -- mempty is the identity of the monoid m
    Const x <*> Const y = Const (x <> y) -- <> is the binary operation of the monoid m
\end{minted}
% \vspace{-3mm}
\caption{Standard library definitions.}\label{fig-stdlib}
% \vspace{-4mm}
\end{figure}

@@

\noindent
The big difference compared to \hs{sprsh1} is that the computation now takes
place in a \hs{Monad}, which allows us to extract the value of \hs{c1} and
\hs{fetch} \emph{different keys depending on whether or not \hs{c1}~\hs{==}~\hs{1}}.

@@

\begin{minted}[xleftmargin=5pt]{haskell}
compute :: Task Monad k v -> Store i k v -> v
compute task store = runIdentity $ run task (\k -> Identity@\,@(getValue k store))
\end{minted}

\noindent
Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-stdlib}).
The use of \hs{Identity} just fixes the `impedance mismatch' between the
function \hs{getValue}, which returns a pure value~\hs{v}, and the \hs{fetch}
argument of the \hs{task}, which must return an \hs{f}~\hs{v} for some \hs{f}.
To fix the mismatch, we wrap the result of \hs{getValue} in the \hs{Identity}
monad: the function
\hs{\@@k}~\hs{->}~\hs{Identity}~\hs{(@@getValue}~\hs{k}~\hs{store)} has the type
\hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a \hs{task}. The
result has type \hs{Identity}~\hs{v}, which we unwrap with \hs{runIdentity}.

@@

The presented definition of correctness needs to be adjusted to accommodate
non-deterministic tasks and shallow cloud builds, as will be discussed
in sections~\S\ref{sec-non-determinism} and~\S\ref{sec-cloud-aspects},
respectively.

@@

\noindent
The main case ($m>0 \wedge n>0$) starts by looking up the pair of indices
\hs{(}\hs{m,}~\hs{n)} in the \hs{cache}. If the cache has \hs{Nothing}, we
calculate the resulting \hs{value} as before and \hs{modify} the cache
accordingly; otherwise, if we have a cache hit, we return the obtained value
immediately, skipping the actual calculation and thus potentially saving a large
amount of work. Indeed you do not want to recompute $A(4,2)=2^{65536}-3$
unnecessarily; all of its 19,729 decimal digits have already been helpfully
computed, e.g. see~\cite{ackermann42}. We will use this technique in our models
of cloud build systems in~\S\ref{sec-implementations}.

>>>>

This section presents abstractions that allow us to express
all the intricacies of build systems discussed in~\S\ref{sec-background},
and express complex build systems from simple primitives. Specifically, we
present the \emph{task} and \emph{build} abstractions in~\S\ref{sec-task}
and~\S\ref{sec-general-build}, respectively. Sections~\S\ref{sec-scheduler},
\S\ref{sec-rebuilder} and~\S\ref{sec-implementations} scrutinise the
abstractions further and provide concrete implementations for several build
systems.

@@

We model build systems concretely, as Haskell programs. To that end,
Fig.~\ref{fig-types} provides the type signatures for all key abstractions
introduced in the paper. For example, \hs{Store}~\hs{i}~\hs{k}~\hs{v} is the
type of stores, with several associated functions (\hs{getValue}, etc.). We use
\hs{k} as a type variable ranging over keys, \hs{v} for values, and \hs{i} for
the persistent build information. Fig.~\ref{fig-stdlib} lists standard library
definitions.

@@

\noindent
Here \hs{c} stands for \emph{constraint}, such as \hs{Applicative}
(\S\ref{sec-why-polymorphism} explains why we need it).
A \hs{Task} describes a single build task, while \hs{Tasks} associates a
\hs{Task} to every non-input key; input keys are associated with \hs{Nothing}.
The highly-abstracted type \hs{Task} describes how to build a value given
a way to build its dependencies, and is best explained by an example.
Consider this \Excel spreadsheet:
% \footnote{Readers familiar with \emph{lenses} or \emph{profunctor optics} might
% recognise a familiar pattern. We discuss this in~\S\ref{sec-related-optics}.}

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Functor     f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard State monad from Control.Monad.State
data State s a
instance Monad (State s)
get       :: State s s
gets      :: (s -> a) -> State s a
put       :: s -> State s ()
modify    :: (s -> s) -> State s ()
runState  :: State s a -> s -> (a, s)
execState :: State s a -> s -> s
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard types from Data.Functor.Identity and Data.Functor.Const
newtype Identity a = Identity { runIdentity :: a }
newtype Const m a  = Const    { getConst    :: m }
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
instance Functor (Const m) where
    fmap _ (Const m) = Const m
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
instance Monoid m => Applicative (Const m) where
    pure _              = Const mempty   -- mempty is identity for monoid m
    Const x <*> Const y = Const (x <> y) -- <> is the binary operation for m
\end{minted}
% \vspace{-3mm}
\caption{Standard library definitions.}\label{fig-stdlib}
% \vspace{-4mm}
\end{figure}

@@

\noindent
The big difference compared to \hs{sprsh1} is that the computation now takes
place in a \hs{Monad}, which allows us to extract the value of \hs{c1} and
\hs{fetch} \emph{different keys} depending on whether or not \hs{c1}~\hs{==}~\hs{1}.

@@

\begin{minted}[xleftmargin=5pt]{haskell}
compute :: Task Monad k v -> Store i k v -> v
compute task store = runIdentity $ run task $
    \k -> Identity $ getValue k store
\end{minted}

\noindent
Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-stdlib}).
The use of \hs{Identity} just fixes the `impedance mismatch' between the
function \hs{getValue}, which returns a pure value~\hs{v}, and the \hs{fetch}
argument of the \hs{task}, which must return an \hs{f}~\hs{v} for some \hs{f}.
To fix the mismatch, we wrap the result of \hs{getValue} in the \hs{Identity}
monad: the function
\hs{\@@k}~\hs{->}~\hs{Identity}~\hs{\$}~\hs{getValue}~\hs{k}~\hs{store} has the type
\hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a \hs{task}. The
result has type \hs{Identity}~\hs{v}, which we unwrap with \hs{runIdentity}.

@@

The presented definition of correctness needs to be adjusted for
non-deterministic tasks and shallow cloud builds, as will be discussed
in sections~\S\ref{sec-non-determinism} and~\S\ref{sec-cloud-aspects},
respectively.

@@

\noindent
The main case ($m>0 \wedge n>0$) starts by looking up the pair of indices
\hs{(}\hs{m,}~\hs{n)} in the \hs{cache}. If the cache has \hs{Nothing}, we
calculate the resulting \hs{value} as before and \hs{modify} the cache
accordingly; otherwise, if we have a cache hit, we return the obtained value
immediately, skipping the actual calculation and thus potentially saving a large
amount of work. Indeed you do not want to recompute $A(4,2)=2^{65536}-3$
unnecessarily; all of its 19,729 decimal digits have already been helpfully
computed, e.g. see~\cite{ackermann42}. We will use \hs{MonadState} tasks in our
models of cloud build systems in~\S\ref{sec-implementations}.
